\label{chap:methodology}

This chapter presents the methodology of our integrated approach, which we term DDECCS (Deep Descriptive Clustering with Consensus Representations). The approach combines pretrained deep feature extraction with consensus-based ensemble clustering and interpretable cluster descriptions via Integer Linear Programming (ILP). The pipeline operates in four stages---feature extraction, clustering, description generation, and evaluation---which we present in turn, before situating the design relative to DECCS and DDC and justifying its main choices. Unlike approaches that jointly train feature representations and clustering objectives---which can suffer from loss-balancing instability, as we discuss in Section~\ref{sec:design_rationale}---our method decouples feature extraction from clustering. This leverages the strong visual representations already learned by modern pretrained networks while allowing each downstream component to operate independently and reliably.

\paragraph{Terminology.} The DDC paper uses the term ``tags'' to refer to semantic properties of instances (e.g., ``furry,'' ``has wheels''). The dataset literature (AwA2, aPY) uses ``attributes'' for the same concept. Throughout this thesis, we use ``attributes'' as the primary term and note that DDC's ``tags'' are synonymous.

\section{Problem Definition}
\label{sec:problem_def}

Given a dataset $\mathcal{D} = \{(\mathbf{x}_i, \mathbf{t}_i)\}_{i=1}^{N}$, where $\mathbf{x}_i$ is an image and $\mathbf{t}_i \in [0,1]^M$ is its associated semantic attribute vector over $M$ attributes, our goal is to produce two outputs:

\begin{enumerate}
    \item A partition of the data into $K$ clusters $\{C_1, \ldots, C_K\}$ that aligns with the underlying semantic categories.
    \item For each cluster $C_k$, a concise description $D_k \subset \{1, \ldots, M\}$: a small subset of attributes that characterizes the cluster. The descriptions are \emph{orthogonal}, meaning the ILP discourages any attribute from being reused across many clusters, so that descriptions remain distinctive across clusters.
\end{enumerate}

The clustering is assessed by its agreement with ground-truth class labels, and the descriptions by how well they cover their clusters and how distinctive they are; the specific metrics are defined in Chapter~\ref{chap:experiments}.

\section{Pipeline Architecture}

Our pipeline consists of four stages, illustrated in Figure~\ref{fig:pipeline}:

\begin{enumerate}
    \item \textbf{Feature Extraction}: Frozen ResNet-101 pretrained on ImageNet extracts 2048-dimensional feature vectors from input images.
    \item \textbf{Clustering}: Either K-means (baseline) or DECCS consensus ensemble assigns images to clusters.
    \item \textbf{Description Generation}: DDC's ILP formulation generates concise, orthogonal cluster descriptions using semantic attributes.
    \item \textbf{Evaluation}: Standard clustering metrics (NMI, ACC, ARI, Silhouette) and interpretability metrics (TC, ITF) assess the results.
\end{enumerate}

\begin{figure}[H]
    \centering
    \begin{tikzpicture}[
        node distance=1.8cm,
        block/.style={rectangle, draw, fill=blue!20, text width=3.5cm, text centered, rounded corners, minimum height=1cm},
        arrow/.style={->, >=stealth, thick}
    ]
        \node (input) [block] {Input Images};
        \node (resnet) [block, below of=input] {ResNet-101\\(frozen, pretrained)};
        \node (features) [block, below of=resnet] {2048-dim Features};
        \node (kmeans) [block, below left of=features, xshift=-2.5cm, yshift=-0.5cm] {K-means};
        \node (deccs) [block, below right of=features, xshift=2.5cm, yshift=-0.5cm] {DECCS Consensus\\(KMeans, Spectral,\\GMM, Agglomerative)};
        \node (clusters) [block, below of=features, yshift=-2.5cm] {Cluster Assignments};
        \node (ilp) [block, below left of=clusters, xshift=-2.5cm, yshift=-0.5cm] {ILP Description\\(DDC Eq.\ 2--4)};
        \node (eval) [block, below right of=clusters, xshift=2.5cm, yshift=-0.5cm] {Evaluation\\(NMI, ACC, TC, ITF)};
        \node (tags) [block, left of=ilp, xshift=-2.5cm] {Semantic\\Attributes};

        \draw[arrow] (input) -- (resnet);
        \draw[arrow] (resnet) -- (features);
        \draw[arrow] (features) -- (kmeans);
        \draw[arrow] (features) -- (deccs);
        \draw[arrow] (kmeans) -- (clusters);
        \draw[arrow] (deccs) -- (clusters);
        \draw[arrow] (clusters) -- (ilp);
        \draw[arrow] (clusters) -- (eval);
        \draw[arrow] (tags) -- (ilp);
    \end{tikzpicture}
    \caption{DDECCS pipeline architecture. Images are processed through a frozen ResNet-101 backbone to extract features, which are then clustered using either K-means or DECCS consensus. Cluster descriptions are generated via ILP using semantic attributes.}
    \label{fig:pipeline}
\end{figure}

Algorithm~\ref{alg:ddeccs} summarizes the complete DDECCS pipeline.

\begin{algorithm}[H]
\caption{DDECCS Pipeline}
\label{alg:ddeccs}
\begin{algorithmic}[1]
    \Require Image dataset $\mathcal{X}$, attribute matrix $T$, number of clusters $K$, mode $\in \{\texttt{kmeans}, \texttt{deccs}, \texttt{ddc}, \texttt{ddeccs}\}$
    \Ensure Cluster assignments $\mathbf{y}$, cluster descriptions $\mathbf{D}$ (if ILP mode)

    \State \textbf{Step 1: Feature Extraction}
    \State Load frozen ResNet-101 (pretrained on ImageNet, final layer removed)
    \State $F \gets \emptyset$
    \For{each image $x_i \in \mathcal{X}$}
        \State Resize $x_i$ to $224 \times 224$, normalize with ImageNet mean/std
        \State $\mathbf{f}_i \gets \text{ResNet-101}(x_i) \in \mathbb{R}^{2048}$ \Comment{Global average pooling output}
        \State $F \gets F \cup \{\mathbf{f}_i\}$
    \EndFor
    \State Cache $F$ to disk for subsequent runs

    \State
    \State \textbf{Step 2: Clustering}
    \State Standardize $F$ to zero mean, unit variance
    \If{mode $\in \{\texttt{kmeans}, \texttt{ddc}\}$}
        \State $\mathbf{y} \gets \text{KMeans}(F, K, \text{n\_init}=10)$
    \ElsIf{mode $\in \{\texttt{deccs}, \texttt{ddeccs}\}$}
    \State $\pi_1 \gets \text{KMeans}(F, K)$
    \State $\pi_2 \gets \text{SpectralClustering}(F, K, k\text{-NN affinity}, k=15)$
    \State $F_{\text{pca}} \gets \text{PCA}(F, d=\min(256, \dim, \max(64, N/2K)))$ \Comment{For GMM numerical stability}
    \State $\pi_3 \gets \text{GMM}(F_{\text{pca}}, K, \text{full covariance})$
    \State $\pi_4 \gets \text{Agglomerative}(F, K, \text{Ward linkage})$
    \State Build co-association matrix: $A_{ij} \gets \frac{1}{4}\sum_{m=1}^{4} \mathbb{1}[\pi_m(i) = \pi_m(j)]$
    \State Sparsify $A$ using $k$-nearest neighbors ($k=20$)
    \State $\mathbf{y} \gets \text{SpectralClustering}(A, K)$
    \EndIf

    \State
    \State \textbf{Step 3: ILP Description Generation} (only if mode $\in \{\texttt{ddc}, \texttt{ddeccs}\}$)
    \State Compute adaptive $\alpha \gets \text{clip}(\text{round}(16 \cdot \bar{d}), 2, 8)$ where $\bar{d}$ is mean attribute density
    \For{each cluster $i = 1, \ldots, K$}
        \For{each attribute $j = 1, \ldots, M$}
            \State $Q_{ij} \gets$ mean of attribute $j$ across instances in cluster $i$
        \EndFor
    \EndFor
    \State Solve ILP: minimize $\sum_{i,j} W_{ij}$ subject to:
    \State \quad Coverage: $\sum_j W_{ij} \cdot Q_{ij} \geq \alpha \quad \forall i$
    \State \quad Orthogonality: $\sum_i W_{ij} \cdot Q_{ij} \leq \beta \quad \forall j$ (smallest feasible $\beta$)
    \State \quad $W_{ij} \in \{0, 1\}$
    \State $\mathbf{D} \gets \{(i, \{j : W_{ij} = 1\}) \mid \forall i\}$

    \State
    \State \textbf{Step 4: Evaluation}
    \State Compute NMI, ACC (Hungarian), ARI, Silhouette against ground-truth labels
    \State Compute TC, ITF for each cluster description in $\mathbf{D}$
\end{algorithmic}
\end{algorithm}

\section{Feature Extraction}
\label{sec:feature_extraction}

We use ResNet-101~\citep{He2016} pretrained on ImageNet~\citep{Deng2009} as our feature extractor. The final classification layer is removed, and the output of the global average pooling layer provides a 2048-dimensional feature vector for each image. All backbone weights are frozen during the entire pipeline---no fine-tuning is performed.

Images are preprocessed with standard ImageNet normalization: resized to $224 \times 224$ pixels, converted to tensors, and normalized with mean $[0.485, 0.456, 0.406]$ and standard deviation $[0.229, 0.224, 0.225]$. Features are extracted once and cached to disk for reuse across experiments.

This approach follows DDC~\citep{Zhang2021}, which uses the same frozen, pre-trained ResNet-101 features---DDC additionally trains a small encoder on top of them, whereas we cluster the features directly. The use of pretrained features is well-established in the clustering literature; K-means on these frozen ResNet-101 features already achieves NMI = 0.869 on AwA2, demonstrating that the backbone captures sufficient class-discriminative information without task-specific training.

% Source: extract_resnet_features() in main_experiments.py

\section{Clustering Methods}
\label{sec:clustering_methods}

We evaluate four clustering configurations that systematically vary the clustering method and the presence of interpretable descriptions:

\begin{table}[H]
    \centering
    \begin{tabular}{lll}
        \hline
        \textbf{Mode} & \textbf{Clustering} & \textbf{Interpretability} \\
        \hline
        \texttt{kmeans} & K-means & --- \\
        \texttt{deccs} & DECCS consensus & --- \\
        \texttt{ddc} & K-means + ILP & Cluster descriptions \\
        \texttt{ddeccs} & DECCS consensus + ILP & Cluster descriptions \\
        \hline
    \end{tabular}
    \caption{Four experimental modes. DDECCS (bottom row) is the thesis contribution, combining consensus clustering with interpretable descriptions. The clustering assignments of \texttt{ddc} are identical to \texttt{kmeans}, and those of \texttt{ddeccs} are identical to \texttt{deccs}; the modes differ only in whether the post-hoc ILP description step is applied.}
    \label{tab:modes}
\end{table}

\paragraph{A note on the \texttt{ddc} and \texttt{ddeccs} mode names.} The \texttt{ddc} mode is \emph{not} a reimplementation of the original Deep Descriptive Clustering method of Zhang and Davidson~\citep{Zhang2021}. The original DDC trains a small encoder (three fully-connected layers) on top of frozen ResNet-101 features, maximizing mutual information and using a self-generated pairwise loss, and solves the ILP \emph{iteratively inside} the training loop so that the explanation can reshape the clustering. In contrast, our \texttt{ddc} mode performs no training: it applies the ILP \emph{once, post-hoc}, to the K-means clustering of frozen features. We use the name only because the mode pairs a single-algorithm clustering with ILP-based descriptions, mirroring DDC's combination at a high level. The \texttt{deccs} and \texttt{ddeccs} modes are named analogously: as detailed in Section~\ref{sec:deccs_consensus}, our consensus step adapts the consensus idea of DECCS in a post-hoc fashion rather than reimplementing its trained representation learning. We do not have a working reimplementation of the original DDC, and all comparisons to it in Chapter~\ref{chap:experiments} use its published results.

\subsection{K-means Baseline}

K-means clustering~\citep{Lloyd1982} is applied directly to the 2048-dimensional ResNet features with $K$ clusters, 10 random initializations, and the Elkan algorithm for efficiency. This serves as the baseline against which consensus clustering is compared.

% Source: KMeans(n_clusters=K, random_state=seed, n_init=10) in main_experiments.py

\subsection{DECCS Consensus Clustering}
\label{sec:deccs_consensus}

DECCS~\citep{Miklautz2022} improves clustering robustness through ensemble consensus. Our implementation builds a consensus from four base clustering algorithms applied to the standardized features. Each algorithm makes a different assumption about cluster shape---its \emph{inductive bias}---and these assumptions are complementary, which is precisely what makes the consensus robust: a grouping that several algorithms with different biases agree on is unlikely to be an artifact of any single one.

\begin{enumerate}
    \item \textbf{K-means} assigns each point to the nearest of $K$ centroids and minimizes within-cluster variance. It implicitly assumes clusters are convex, roughly spherical, and of similar size, and is fast but fails on elongated or non-convex clusters. We use the Elkan algorithm for efficiency.
    \item \textbf{Spectral Clustering} builds a $k$-nearest-neighbour affinity graph ($k = 15$) and clusters using the eigenvectors of the graph Laplacian. Because it operates on graph connectivity rather than raw distances, it recovers non-convex, manifold-shaped clusters that K-means cannot.
    \item \textbf{Gaussian Mixture Model (GMM)} models the data as a mixture of $K$ Gaussians with full covariance, giving each point a soft probabilistic membership and capturing elliptical clusters of differing orientation and size. Estimating a full covariance matrix in 2048 dimensions is numerically unstable, so we first reduce the features with PCA before fitting the GMM.
    \item \textbf{Agglomerative Clustering} starts with each point as its own cluster and repeatedly merges the pair whose merger least increases within-cluster variance (Ward linkage). It is hierarchical and makes no centroid or distributional assumption, capturing nested structure.
\end{enumerate}

The ensemble results are combined via a sparse co-association matrix $A$, where $A_{ij}$ counts how often instances $i$ and $j$ are assigned to the same cluster across the base clusterers, restricted to $k$-nearest neighbours for efficiency~\citep{Fred2005}. The final clustering applies Spectral Clustering to this consensus matrix.

We apply Spectral Clustering rather than K-means to this matrix because the co-association values define a graph---instances are nodes and co-association frequencies are edge weights---and Spectral Clustering is designed to partition exactly this kind of graph structure; running K-means on the matrix would instead treat the co-association values as raw features and discard the graph interpretation~\citep{Fred2005}.

This differs from the original DECCS, which trains an autoencoder with a consensus loss during representation learning. Our post-hoc consensus applies the ensemble after feature extraction, trading the iterative refinement of learned representations for the reliability and reproducibility of frozen pretrained features. As our experiments demonstrate (Section~\ref{sec:results}), this trade-off is favorable: DECCS consensus provides significant improvements on datasets where individual clustering methods are less reliable (e.g., +14.4\% relative NMI on aPY's 32-class task).

% Source: get_base_clusterings() and build_sparse_consensus() in utils.py

\section{Cluster Description via Integer Linear Programming}
\label{sec:ilp}

Following DDC~\citep{Zhang2021}, we generate interpretable cluster descriptions by solving an Integer Linear Programming (ILP) problem that selects a small, orthogonal set of semantic attributes for each cluster.

\subsection{Problem Formulation}

Given cluster assignments $\{C_1, \ldots, C_K\}$ and a set of $M$ semantic attributes, we define the attribute coverage matrix $Q \in \mathbb{R}^{K \times M}$ where $Q_{ij}$ is the mean value of attribute $j$ across all instances in cluster $C_i$. The ILP solves for a binary allocation matrix $W \in \{0,1\}^{K \times M}$ where $W_{ij} = 1$ indicates that cluster $C_i$ is described by attribute $j$.

\textbf{Objective} (DDC Eq.\ 2): Minimize the total number of attributes used:
\begin{equation}
    \arg\min_W \sum_{i,j} W_{ij}
    \label{eq:ilp_obj}
\end{equation}

\textbf{Coverage constraint} (DDC Eq.\ 3): Each cluster must have sufficient attribute coverage:
\begin{equation}
    \sum_{j=1}^{M} W_{ij} \cdot Q_{ij} \geq \alpha \quad \forall i \in \{1, \ldots, K\}
    \label{eq:ilp_coverage}
\end{equation}

\textbf{Orthogonality constraint} (DDC Eq.\ 4): the total coverage-weighted usage of each attribute across all clusters is bounded by $\beta$, which keeps descriptions distinctive by discouraging any single attribute from being shared widely across clusters:
\begin{equation}
    \sum_{i=1}^{K} W_{ij} \cdot Q_{ij} \leq \beta \quad \forall j \in \{1, \ldots, M\}
    \label{eq:ilp_orthogonal}
\end{equation}

The solver searches for the smallest feasible $\beta$ starting from $\beta = 1$, ensuring the tightest possible orthogonality. We use the PuLP library~\citep{Mitchell2011} with the CBC solver and a 30-second timeout per $\beta$ value.

\subsection{Adaptive \texorpdfstring{$\alpha$}{alpha} Parameter}
\label{sec:adaptive_alpha}

DDC uses a fixed $\alpha = 8$, which assumes dense attribute annotations. For datasets with sparse attributes, $\alpha = 8$ can make the ILP infeasible, and even when it is feasible it forces overly long descriptions. We therefore introduce an adaptive $\alpha$ that scales with the mean attribute value, and explain the issue with a concrete example.

The coverage constraint (Eq.~\ref{eq:ilp_coverage}) requires $\sum_j W_{ij} \cdot Q_{ij} \geq \alpha$ for each cluster, where the $Q_{ij}$ are the mean attribute values within the cluster. A cluster's maximum achievable coverage is therefore $\sum_j Q_{ij}$, which is small when the attributes are sparse.

On aPY, attributes are binary and sparse, with mean value $\bar{d} \approx 0.087$ on the 15-class subset. A typical cluster might have $Q$ values like Wheel = 0.30, Metal = 0.25, Pedal = 0.15, Furry = 0.02---most values are small because only $8.7\%$ of attribute entries are 1. Even selecting all 64 attributes yields a coverage of only about $64 \times 0.087 \approx 5.6$ for a typical cluster---well below $8$---so $\alpha = 8$ is infeasible on aPY.

On AwA2, attributes are continuous (normalized to $[0,1]$) with mean value $\bar{d} \approx 0.217$. The dominant attributes in a clean cluster have high coverage---for a tiger cluster, \textit{stripes}, \textit{orange}, and \textit{muscle} are each around $0.85$--$0.9$---so four such attributes already sum to roughly $3.4$. Here $\alpha = 8$ would be feasible but would force each cluster to include ten or more attributes to accumulate that much coverage, producing verbose, less distinctive descriptions.

Our adaptive formula scales $\alpha$ proportionally to $\bar{d}$:

\begin{equation}
    \alpha = \text{clip}\left(\text{round}(16 \cdot \bar{d}), \; 2, \; 8\right)
    \label{eq:alpha_adaptive}
\end{equation}

where $\bar{d}$ is the mean attribute value across all instances and all attributes (i.e., the average entry in the attribute matrix); for binary attributes this equals the fraction of non-zero entries. The constant 16 recovers DDC's $\alpha = 8$ at a reference density of $\bar{d} = 0.5$ and scales the requirement down for sparser attributes. The clip to $[2, 8]$ guarantees at least two meaningful attributes per cluster (lower bound) and never exceeds DDC's original setting (upper bound). This yields $\alpha = 3$ for AwA2 ($\bar{d} = 0.217$) and $\alpha = 2$ for aPY (both the 15-class subset, $\bar{d} = 0.087$, and the full 32-class set, $\bar{d} = 0.111$), producing concise descriptions of roughly four to six attributes per cluster on both datasets.

% Source: solve_ilp_descriptions() in main_experiments.py

\section{Design Rationale: Why Frozen Features Instead of a Trained Representation}
\label{sec:design_rationale}

Our initial approach attempted to replicate DDC's trained-encoder approach---a trainable MLP on frozen ResNet features, optimized with mutual information maximization and self-generated pairwise constraints. This did not scale: the trained representation collapsed on the full datasets, performing far worse than K-means on the same frozen features. We give the full diagnosis of this failure in Sections~\ref{sec:early_experiments} and~\ref{sec:why_failed}.

This outcome motivates the design above. Rather than training features that already separate the classes well, we focus the contribution on two decoupled components: consensus clustering for robustness, and post-hoc ILP descriptions for interpretability. Decoupling them lets us evaluate clustering quality and description quality independently and yields a deterministic, reproducible pipeline free of training instabilities---at the cost of task-specific feature adaptation, a trade-off we examine in Chapter~\ref{chap:discussion}.